---
title: "tutorial"
output: html_document
---

# Seurat Basic Workflow

```{r}
library(Seurat)
library(tidyverse)
```

## Loading in Data

We got our data from 10x genomics, [Human PBMCs Labelled With Extracellular and Intracellular Antibodies Using the Proteintech MultiPro™ Human Fixed Cell Immune Profiling Antibody Cocktail](https://www.10xgenomics.com/datasets/80k_Human_PBMCs_PTG_MultiproPanel_IC_4plex). From here, we downloaded "[Gene Expression - Feature / cell matrix HDF5 (per-sample)](https://cf.10xgenomics.com/samples/cell-exp/9.0.0/80k_Human_PBMCs_PTG_MultiproPanel_IC_4plex_PBMC_BC1_AB1_Rested_Unstained/80k_Human_PBMCs_PTG_MultiproPanel_IC_4plex_PBMC_BC1_AB1_Rested_Unstained_count_sample_filtered_feature_bc_matrix.h5)".

The first step is to load in data, which Seurat handles.

```{r}
sparse <- Read10X_h5(filename = "80k_Human_PBMCs_PTG_MultiproPanel_IC_4plex_PBMC_BC1_AB1_Rested_Unstained_count_sample_filtered_feature_bc_matrix.h5")
```

This has multiple modalities, so we can use a str() to observe

```{r}
str(sparse)
```

We just want gene expression, so let's select it

```{r}
counts <- sparse$`Gene Expression`
counts
```

We can now initialize the Seurat object

```{r}
sparse <- CreateSeuratObject(counts = counts, project = "PBMC", min.cells = 3, min.features = 200)
str(sparse)
```

## QC

QC is always crucial!

### MT Reads

We basically want to filter out cells which have really low number of counts. This is indicative of a bad read because the cell was not read fully or the gene died (due to higher mitochondrial stress). We also want to filter out high number of counts because that could mean that the cell was read 2x or was read alongside another cell (aka made a doublet).

```{r}
view(sparse@meta.data)
```

```{r}
sparse[["percent.mt"]] <- PercentageFeatureSet(sparse, pattern = "^MT-") # save percentage of mitochondrial reads as another variable in metadata
```

We can also visualize our data using VlnPlot, a built-in function from Seurat. We then use a FeatureScatter to create a scatter plot of nCount and nFeatures. Hypothetically, cells with a high count should also have a high number of genes (features).

```{r}
VlnPlot(sparse, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol=3)
FeatureScatter(sparse, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  geom_smooth(method = 'lm') # line of best fit
```

This data looks pretty good, but if we see something in the bottom right or or top left that doesn't fit this log curve, you should probably filter it out.

### Filtering

We now filter out our "bad" reads. In this instance, we only keep cells with more than 200 but less than 2500 genes and cells with less than 5% mito reads.

```{r}
sparse <- subset(sparse, subset = nFeature_RNA > 200 & nFeature_RNA <  2500 & percent.mt < 5)
```

We're not going to do it here, but we can use the "doublet finder" package to find doublets. In this tutorial, we will just filter out based on percent mito reads.

## Normalize Data

Seurat has a built-in function to normalize data.

```{r}
sparse <- NormalizeData(sparse)
```

### Observing Commands Using @commands

We can observe all of the post processing that we've done on the Seurat object by looking for the @commands in the Seurat object.

```{r}
str(sparse)
```

## Identify Highly Variable Features

We want to identify highly variable features because it has been found that focusing on just these features shows a higher degree of significance at the end. We can use the FindVariableFeatures function built into Seurat.

```{r}
sparse <- FindVariableFeatures(sparse, selection.method = "vst", nfeatures = 2000) # default settings
top10 <- head(VariableFeatures(sparse), 10) # VariableFeatures lets us observe the top variable features
```

We can now visualize these features using VariableFeaturePlot

```{r}
plot1 <- VariableFeaturePlot(sparse)
LabelPoints(plot = plot1, points = top10, repel = T)
```

We can now see the most variable genes.

## Scaling

In sc dataset, there is lots of inherent noise from stages in cell cycle, batch noise, etc. We want to make sure that clustering downstream is because of gene expression and not any of the aforementioned effects. To prevent this, we do scaling.

```{r}
all.genes <- rownames(sparse)
sparse <- ScaleData(sparse, features = all.genes)
```

Let's do another check on our Seurat object.

```{r}
str(sparse)
```

Just as a review, **under RNA Assay we have @ counts which is the raw sparse matrix, @ data which tells us the log normalized counts, and @ scale.data is the same as data but scaled** as in this step.

## Linear Dimensionality Reduction

We can now run PCA.

```{r}
sparse <- RunPCA(sparse, features = VariableFeatures(object = sparse))
```

Let's visualize the principle components.

```{r}
print(sparse[["pca"]], dims = 1:5, nfeatures = 5)
DimHeatmap(sparse, dims = 1, cells = 500, balanced = T)
```

This shows us the heterogeneity of the matrix.

We can create an Elbow Plot that tells us the dimensionality of each of the PCs.

```{r}
ElbowPlot(sparse)
```

It looks like we only need to consider about 15 PCs to describe the dataset.

## Clustering

In the previous step, we determined that we need about 15 PCs to describe the dataset, so that is the dimensionality parameter we will use with the Seurat FindNeighbors function for clustering.

```{r}
sparse <- FindNeighbors(sparse, dims = 1:15)
```

We want to tune resolution, which will determine the number of clusters in our analysis.

```{r}
sparse <- FindClusters(sparse, resolution = c(0.01, 0.05, 0.1, 0.3))
# View(sparse@meta.data) # look at the different clusters
```

We want to now see what has the best clustering. To do this, we can use a DimPlot to visualize the results.

```{r}
DimPlot(sparse, group.by = "RNA_snn_res.0.01", label = T)
DimPlot(sparse, group.by = "RNA_snn_res.0.05", label = T)
DimPlot(sparse, group.by = "RNA_snn_res.0.1", label = T)
DimPlot(sparse, group.by = "RNA_snn_res.0.3", label = T)
```

.05 looks the best. Let's assign that as the "correct" clustering.

```{r}
Idents(sparse) <- "RNA_snn_res.0.05"
# Idents(sparse) look at the assignments
```
